Free Agency in Baseball: MLB Teams Aren’t Getting The Player They Bargained For

### 2022-11-14 manuscript/manuscript.tex
final draft check

\section*{Abstract}
Every winter, professional baseball teams bid for the services of the MLB’s most talented hitters. Baseball’s top stars often 
sign contracts for upwards of half a decade, guaranteeing them tens, if not hundreds of millions of dollars. However, some of 
these contracts work out better than others. When teams shell out inordinate amounts of money, they are gambling on their
investment panning out; when it doesn’t, the consequences can be drastic. The conventional wisdom in baseball states that
players who sign contracts with new teams tend to underperform expectations in their new city. Mostly supported by anecdotal 
and circumstantial evidence, this theory can be scientifically put to the test. By comparing player performance from 2021 to
2022 for players that switched baseball clubs this past winter, we can either support or contradict this long-repeated 
conjecture. In this paper we will explore the differences from year to year in numerous key baseball statistics. For each
significant statistical finding, practical explanations will be offered based on existing literature and practical causal 
arguments.

\section*{Keywords}
Baseball Contracts; Baseball Statistics; Free Agency; Injuries; Player Decline; Player Durability; Player Performance

\section{Introduction}
On May 5th, 2022, Boston Red Sox second baseman Trevor Story walked back to the dugout after a strikeout, his fourth such 
result of the night. Boos rained down on him from Red Sox fans who had, to this point, watched their beloved team muster zero 
runs through seven innings. Story had made his Red Sox debut just 27 days ago after signing a six year, 140 million dollar 
contract over the winter with Boston. How did the multimillion dollar experiment go sideways so quickly?

@@

\section{Data}
The data that has been collected for this study came mainly from Baseball Reference and Spotrac, two websites that contain 
nearly every morsel of baseball data one could desire. Baseball Reference mostly provided the true statistics and game-by-game 
breakdowns, while Spotrac mainly provided contract and free agency information.

@@

\section{Methods}
The basic statistical mechanism used in this study was analysis of variance. By utilizing a one-way ANOVA, I was able to 
compare means for all of the aforementioned variables from 2021 to 2022. The one-way ANOVA assumes that the data follows a 
normal distribution, which is an acceptable approximation given the large sample size of players. Each counting statistic, 
value added statistic, and health statistic was run through this one-way analysis of variance test. For each of these 
statistics, the null and research hypotheses followed the same general form. The null hypothesis was that the mean value 
among all 50 players for these statistics in 2021 was equal to the mean value among the same 50 players in 2022. This 
hypothesis is the one we seek to reject, thus we need a research hypothesis that has the burden of proof. We must be able 
to measure the strength of the research hypothesis via our P-value. Therefore, our research (alternate) hypothesis is as 
follows. The mean value among all 50 players for these statistics in 2021 was not equal to the mean value among the same 50 
players in 2022. We are seeking to uncover whether there was significant change in player performance between the two years, 
so these hypotheses will suffice.

@@

\section{Simulation and Application}
The following tables contain the results of the one-way analysis of variance for our counting statistics. The first column 
describes what the statistic is that is being tested. The second column contains the average number of that statistic that 
the 50 players recorded during the 2021 season. The third column contains the average number of that statistic that the 50 
players recorded during the 2022 season. The fourth column contains the raw increase or decrease in that statistic from 2021 
to 2022. The fifth column contains the percentage increase or decrease in that statistic from 2021 to 2022. The sixth column 
contains the P-value for difference in the two means as calculated by SAS 9.4. The final column contains the significance 
marking for the statistic. A statistic with one asterisk indicates rejection of the null hypothesis at alpha = 0.10. A 
statistic with two asterisks indicates rejection of the null hypothesis at alpha = 0.05. A statistic with three asterisks 
indicates rejection of the null hypothesis at alpha = 0.01.

@@

Stat
2021
2022
P-Value
Signif.
AVG.
4391/17591 = .24962
3577/14557 = .24572
0.4237

SLG.
7587/17591 = .43130
5830/14557 = .40049
<0.0001
***
HR%
742/19853 = 3.737%
491/16198 = 3.031%
0.0002
***
BB%
1831/19853 = 9.223%
1310/16198 = 8.087%
0.0001
***
SO%
4413/19853 = 22.228%
3568/16198 = 22.027%
0.6455

SB%
258/332 = 77.711%
184/244 = 75.410%
0.5157

@@

>>>>

\section*{Abstract}

Every winter, professional baseball teams bid for the services of the MLB’s most talented hitters. Baseball’s top stars often 
sign contracts for upwards of half a decade, guaranteeing them tens, if not hundreds of millions of dollars. However, some of 
these contracts work out better than others. When teams shell out inordinate amounts of money, they are gambling on their
investment panning out; when it doesn’t, the consequences can be drastic. The conventional wisdom in baseball states that
players who sign contracts with new teams tend to underperform expectations in their new city. Mostly supported by anecdotal 
and circumstantial evidence, this theory can be scientifically put to the test. By comparing player performance from 2021 to
2022 for players that switched baseball clubs this past winter, we can either support or contradict this long-repeated 
conjecture. In this paper we will explore the differences from year to year in numerous key baseball statistics. For each
significant statistical finding, practical explanations will be offered based on existing literature and practical causal 
arguments.

\section*{Keywords}

Baseball Contracts; Baseball Statistics; Free Agency; Injuries; Player Decline; Player Durability; Player Performance

\section{Introduction}

On May 5th, 2022, Boston Red Sox second baseman Trevor Story walked back to the dugout after a strikeout, his fourth such 
result of the night. Boos rained down on him from Red Sox fans who had, to this point, watched their beloved team muster zero 
runs through seven innings. Story had made his Red Sox debut just 27 days ago after signing a six year, 140 million dollar 
contract over the winter with Boston. How did the multimillion dollar experiment go sideways so quickly?

@@

\section{Data}

The data that has been collected for this study came mainly from Baseball Reference and Spotrac, two websites that contain 
nearly every morsel of baseball data one could desire. Baseball Reference mostly provided the true statistics and game-by-game 
breakdowns, while Spotrac mainly provided contract and free agency information.

@@

\section{Methods}

The basic statistical mechanism used in this study was analysis of variance. By utilizing a one-way ANOVA, I was able to 
compare means for all of the aforementioned variables from 2021 to 2022. The one-way ANOVA assumes that the data follows a 
normal distribution, which is an acceptable approximation given the large sample size of players. Each counting statistic, 
value added statistic, and health statistic was run through this one-way analysis of variance test. For each of these 
statistics, the null and research hypotheses followed the same general form. The null hypothesis was that the mean value 
among all 50 players for these statistics in 2021 was equal to the mean value among the same 50 players in 2022. This 
hypothesis is the one we seek to reject, thus we need a research hypothesis that has the burden of proof. We must be able 
to measure the strength of the research hypothesis via our P-value. Therefore, our research (alternate) hypothesis is as 
follows. The mean value among all 50 players for these statistics in 2021 was not equal to the mean value among the same 50 
players in 2022. We are seeking to uncover whether there was significant change in player performance between the two years, 
so these hypotheses will suffice.

@@

\section{Simulation and Application}

The following tables contain the results of the one-way analysis of variance for our counting statistics. The first column 
describes what the statistic is that is being tested. The second column contains the average number of that statistic that 
the 50 players recorded during the 2021 season. The third column contains the average number of that statistic that the 50 
players recorded during the 2022 season. The fourth column contains the raw increase or decrease in that statistic from 2021 
to 2022. The fifth column contains the percentage increase or decrease in that statistic from 2021 to 2022. The sixth column 
contains the P-value for difference in the two means as calculated by SAS 9.4. The final column contains the significance 
marking for the statistic. A statistic with one asterisk indicates rejection of the null hypothesis at alpha = 0.10. A 
statistic with two asterisks indicates rejection of the null hypothesis at alpha = 0.05. A statistic with three asterisks 
indicates rejection of the null hypothesis at alpha = 0.01.

@@

\begin{table}[h!]
    \begin{center}
      \caption{Proportion Statistics}
      \label{tab:table4}
      \begin{tabular}
        \textbf{Stat} & \textbf{2021} & \textbf{2022} & \textbf{P-Value} & \textbf{Signif.}
        \hline
        AVG. & 4391/17591 = .24962 & 3577/14557 = .24572 & 0.4237 &  \\
        SLG. & 7587/17591 = .43130 & 5830/14557 = .40049 & <0.0001 & *** \\
        HR PCT & 742/19853 = 3.737 & 491/16198 = 3.031 & 0.0002 & *** \\
        BB PCT & 1831/19853 = 9.223 & 1310/16198 = 8.087 & 0.0001 & *** \\
        SO PCT & 4413/19853 = 22.228 & 3568/16198 = 22.027 & 0.6455 &  \\
        SB PCT & 258/332 = 77.711 & 184/244 = 75.410 & 0.5157 &  \\
      \end{tabular}
    \end{center}
\end{table}

@@

\section{Discussion}

Our first table showed statistically significant declines in runs, home runs, runs batted in, and total bases. These categories 
share a sort of commonality in that they are all statistics that rely on a hitter being able to have lots of swing power. Home 
runs are almost exclusively considered a sign of power, and our players hit, on average, five less home runs in 2022 than they 
did in 2021. There could be several possible explanations for this finding. A baseball team plays half of its games at its home 
field; when a player switches teams, they are invariably switching the venue in which they play a large chunk of their games. 
Ballpark dimensions vary from stadium to stadium, with the fences differing in height, distance from home plate, and angle. A 
good example of this comes from Trevor Story, one of our 50 sampled players and the one from the anecdote at the beginning of 
my introduction. Coors Field, his old field in Denver, measures 390 feet from home plate to the center field wall. Fenway Park, 
his new field in Boston, measures 420 feet to center field- a whopping 30 foot difference. Hitting a baseball 420 feet is a 
tall task. Players who have played in Boston for years will tell you they never try to hit a ball to center field, lest they 
waste their power trying to get the ball over one of the farthest walls in the entire league. Conversely, Story, playing his 
first season in Boston, is likely unfamiliar with the quirks and oddities that come with being a batter in a particular ballpark 
with unique dimensions. Our sample of 50 players all had to adapt to hitting in a new ballpark, forcing them to restructure 
their entire approach at the plate. This seismic disruption in their play style would also explain why these players were 
slightly less valuable in 2022 than in 2021, as shown in our second table.

I would be remiss if I didn’t also mention that in general, moving teams is as massive a change off the field as it is on the 
field. Anybody who has transplanted themselves into a new city will agree that a massive adjustment period is necessary in 
order to get one’s feet underneath them in a new environment. Shortstop Andrelton Simmons, one of our sample subjects, made 
the transition from small-market Minneapolis to the Windy City of Chicago, the third-largest media market in the United States. 
The aforementioned Story went from Denver, a city over 5,000 feet above sea level, to Boston, which is only a hair over 100 
feet above sea level. It is not difficult to fathom a potential link between player performance issues and the enormous 
life-altering decisions that directly precede them.

Of course, it must be discussed that a sizable portion of the drop-off can be explained by our players’ lack of physical 
fitness in 2022. As could be seen in our third table, game participation fell by nearly 25 games per player from 2021 to 2022. 
This difference was deemed empirically significant, and there is a theoretical explanation for this finding as well. Chapter 
12 of John Cassidy’s book How Markets Fail is entitled “George Akerlof’s Market for Lemons”, wherein he explains that unequal 
access to information plagues almost every sector of our economy. This includes the market for baseball free agents, which 
Cassidy explains suffers from the problem of hidden injury information. He posits that “a player’s original team knows more 
about his health than a potential acquirer, and the decision to release him to free agency may be a sign that it has concerns 
about his health” (Cassidy 154). This hypothesis is consistent with the previously mentioned studies of Garcia et al. which 
found that teams are more willing than ever to part with their own players. While the drop in games played doesn’t necessarily 
reveal a regression in talent or skill, a player needs to be healthy in order to be useful. Our value statistics are tied to 
game participation because a player that is riding the bench with back problems isn’t valuable at all. Teams are, generally, 
required to pay players their massive salaries regardless of how healthy they are, making this trend of bad health in newly 
signed players troubling for big spenders. Teams see big free agency contracts as an investment, and a player can’t provide 
return on that investment unless they’re truly fit to play. It should also be mentioned the effect that switching organizations 
might have on a player’s health. Even if there aren’t any nagging hidden issues like the ones Cassidy hypothesized about, a 
new team means a new training staff and a new set of team doctors. Teams aren’t owed any extensive medical records when they 
sign a player, so medical staff have very little to work from when treating a newly acquired player.

Arguably the most troubling finding for our players, even more so than their dismal health, was their inability to meet 
expectations when they were on the field. Our fourth and final table contained statistical categories whose drop-offs could 
not be explained by not playing as many games in 2022. Every category decreased in some way except for strikeout percentage, 
with slugging percentage, home run percentage, and walk percentage decreasing significantly. It seems as though between power, 
contact, speed, and discipline, power is the performance tool that appears to have evaporated the most in our players. The drop 
in slugging percentage and home run percentage matches up with the drop in other power-related categories from table one. 
Hitting power is a complicated concept. Gratuitous strength doesn’t always correlate with good power, but it certainly does 
play a part. Perfect swing form and unimpeachable timing ability are inarguably more important when it comes to launching the 
baseball- strength means nothing if you cannot get the barrel of the bat on the ball. A baseball swing can be thought of as a 
symphony of moving body parts. If a batter develops even the slightest hitch in their swing, the entire orchestra is thrown out 
of balance and the player stands no chance of hitting the ball squarely. These hitches usually develop from, coincidentally, 
nagging injuries to the back, neck, and extremities. With so many of the players in our sample battling long-term ailments in 
2022, it is not difficult to see how their power was essentially zapped when compared to the season before. These weeks-long 
bouts with injury can cause major performance slumps, draining a player’s confidence and setting them up for failure. Baseball 
is partially a mental game- step up to the plate with a pessimistic mindset, and you’re almost guaranteed to fail. Early-season 
struggles, especially with a brand-new team, can easily spiral into season-long, emotionally torturous campaigns that leave 
fans wondering what exactly the team signed up for when they shelled out millions of dollars.

All of this cynicism is not to say that our players can’t turn their fortunes around. The numbers don’t lie: the players in 
our sample saw a downturn in essentially every single statistical category we decided to investigate. However, we have seen 
players take their lumps with a new team and persevere. Kyle Schwarber, one of the sampled players, saw one of the largest 
year-over-year decreases in batting average from 2021 to 2022. But when his Philadelphia Phillies made their improbable run 
to the World Series this November, Schwarber reversed course and smashed 6 home runs in 17 games, giving Phillies fans hope for 
a brighter campaign ahead. Baseball is, and will continue to be, an endearingly unpredictable sport. Players that switch teams 
will likely always have their growing pains, but the fans that stick it out might just be lucky enough to bear the fruit when 
those players bloom.

\bibliographystyle{chicago}
\bibliography{citations.bib}

\section{Acknowledgements}

Thank you to the University of Connecticut Department of Statistics faculty, and specifically to Dr. Jun Yan, for their 
contribution to and assistance with the writing of this paper.

\section{About The Student Author}

Jason Boghosian is an expected graduate of the University of Connecticut with a B.A. in Statistics and a minor in both 
Public Policy and Urban Studies. He will begin his postgraduate studies in August of 2023 when he pursues a Master of Arts 
in Survey Research at the University of Connecticut.

\section{Press Summary}

The purpose of this study was to examine the effects that switching teams had on professional baseball players in 2022. The 
hope was to investigate whether there were statistical grounds for the claim that players who change organizations generally 
perform worse on their new team. This study focused on the difference from 2021 to 2022 in 18 statistical categories for 50 
sampled MLB players that switched teams between the two seasons. After running statistical tests, the implication of the 
findings was that players’ fitness level, health, and overall hitting performance appeared to drastically regress after 
switching teams.
